Supplementary File 1

### 2019-08-22 fig1_2app.Rmd
article files

---
title: "supplement"
author: "Selcen Ari"
date: "7 Nisan 2019"
output: html_document
---

@@

*Convertion of dataset*: priming_graph function processes the given dataset that includes competing elements in first variable and repressive element in second variable. If the affinity and/or degradation factors are specified in the function, factors are taken into account, are processed with defaults in vice versa. The formulations that are given in equations (1-5) are performed in this function. This step gives the graph object which contains efficiency values of miRNA:competing target pairs in steady-state in terms of amount. It is assumed that the initial target amounts in the dataset is observed after the reppressive activity of miRNAs in steady-state.

![Workflow for simulation of competing endogenous RNA regulations. Graph object in steps 2-6 is saved and updated continuously.](workflow1.jpg){#fig1}

*Transition of variables in graph*: In the previous step, the calculations are performed in the edge variables of the graph object. However, the graph object allows to use node variables, while the node features are handled to the graph. In this direction, update_nodes function carries the amount values to node variables This step must be applied with “once” option because it is primary process.

*Trigger change in graph*: The dataset are assumed as steady-state in previous step and the efficieny coefficients are calculated according to this acceptance. In the network that is found in steady-state conditions, the change is applied to the graph object for disturbtion of steady-state. To provide the disturbtion in the network the workflow offer two methods: update_variables and update_how. The first, a new dataset that is contained competing and repressive element names and current values of these can be processed with update_variables. The second option, the amount of the given node name in update_how function can be changed according to “how” argument.

*Updating current values of variables*: After variables updating in edge varibles, these are carried to node variables. Current and previous values of variables are stored as node variables with update_variables function.

*Simulation of competing behavior of targets*: After the change in the steady-state conditions, the network elements try to gain steady-state again. This process progresses as repeating of regulations after the spreading the changes in the network. In this step, simulation of regulations according to given cycle count in simulate function is applied. After each simulation cycle, the miRNA repression values are re-calculated and the current values of competing elements are found and saved. The process is performed in the edge data and at the same time outputs of the calculations are carried from edge to node data.

@@

```{r, message=FALSE, warning=FALSE}
library(ceRNAnetsim)
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_nodes(once=TRUE)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", title = "Minimal dataset in steady-state conditions")
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_nodes(once=TRUE)%>%
  update_how("Gene2", 2)%>%
  update_nodes()%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors")
```

@@

priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_nodes(once=TRUE)%>%
  update_how("Gene2", 2)%>%
  update_nodes()%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", Downregulation = "blue", title = "Regulations after Gene2 Upregulation", 2)

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_nodes(once=TRUE)%>%
  update_how("Gene2", 2)%>%
  update_nodes()%>%
  simulate(2)%>%
  activate(edges)%>%
  as_tibble()%>%
  select(Competing_name,comp_count_list,effect_list)%>%
  unnest()
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_nodes(once=TRUE)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", title = "Minimal dataset in steady-state conditions")

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_nodes(once=TRUE)%>%
  update_how("Gene2", 2)%>%
  update_nodes()%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors")
```

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_nodes(once=TRUE)%>%
  update_how("Gene2", 2)%>%
  update_nodes()%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors",2)
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_nodes(once=TRUE)%>%
  update_how("Gene2", 2)%>%
  update_nodes()%>%
  simulate(3)%>%
  activate(edges)%>%
  as_tibble()%>%
  select(Competing_name,comp_count_list,effect_list)%>%
  unnest()
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_nodes(once=TRUE)%>%
  update_how("Gene4", 2)%>%
  update_nodes()%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors",2)
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_nodes(once=TRUE)%>%
  update_how("Gene4", 2)%>%
  update_nodes()%>%
  simulate(3)%>%
  activate(edges)%>%
  as_tibble()%>%
  select(Competing_name,comp_count_list,effect_list)%>%
  unnest()
```

>>>>

---
title: "Supplementary File 1"
author: "Selcen Ari"
date: "7 Nisan 2019"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

@@

**Convertion of dataset**: priming_graph function processes the given dataset that includes competing elements in first variable and repressive element in second variable. If the affinity and/or degradation factors are specified in the function, factors are taken into account, are processed with defaults in vice versa. The formulations that are given in equations (1-4) are performed in this function. This step gives the graph object which contains efficiency values of miRNA:competing target pairs in steady-state in terms of amount. It is assumed that the initial target amounts in the dataset is observed after the reppressive activity of miRNAs in steady-state.

![Workflow for simulation of competing endogenous RNA regulations. Graph object in steps 2-6 is saved and updated continuously.](workflow1.png){#fig1}

**Transition of variables in graph**: In the previous step, the calculations are performed in the edge variables of the graph object. However, the graph object allows to use node variables, while the node features are handled to the graph. In this direction, update_nodes function carries the amount values to node variables This step must be applied with “once” option because it is primary process.

**Trigger change in graph**: The dataset are assumed as steady-state in previous step and the efficieny coefficients are calculated according to this acceptance. In the network that is found in steady-state conditions, the change is applied to the graph object for disturbtion of steady-state. To provide the disturbtion in the network the workflow offer two methods: update_variables and update_how. The first, a new dataset that is contained competing and repressive element names and current values of these can be processed with update_variables. The second option, the amount of the given node name in update_how function can be changed according to “how” argument.

**Updating current values of variables**: After variables updating in edge varibles, these are carried to node variables. Current and previous values of variables are stored as node variables with update_variables function.

**Simulation of competing behavior of targets**: After the change in the steady-state conditions, the network elements try to gain steady-state again. This process progresses as repeating of regulations after the spreading the changes in the network. In this step, simulation of regulations according to given cycle count in simulate function is applied. After each simulation cycle, the miRNA repression values are re-calculated and the current values of competing elements are found and saved. The process is performed in the edge data and at the same time outputs of the calculations are carried from edge to node data.

@@

```{r, message=FALSE, warning=FALSE}
#install.packages("devtools")
#devtools::install_github("selcenari/ceRNAnetsim")
library(ceRNAnetsim)
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", title = "Minimal dataset in steady-state conditions")
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_how("Gene2", 2)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors")
```

@@

priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_how("Gene2", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", Downregulation = "blue", title = "Regulations after Gene2 Upregulation", 2)

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression)%>%
  update_how("Gene2", 2)%>%
  simulate(2)%>%
  activate(edges)%>%
  as_tibble()%>%
  select(Competing_name,comp_count_list,effect_list)%>%
  unnest()
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", title = "Minimal dataset in steady-state conditions")

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  vis_graph(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors")
```

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors",2)
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene2", 2)%>%
  simulate(3)%>%
  activate(edges)%>%
  as_tibble()%>%
  select(Competing_name,comp_count_list,effect_list)%>%
  unnest()
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene4", 2)%>%
  simulate_vis(Competing_color = "navajowhite3", mirna_color = "ivory4", Upregulation = "red", title = "Gene2 Upregulation without interaction factors",2)
```

@@

```{r, message=FALSE, warning=FALSE}
priming_graph(minsamp, competing_count = Competing_expression, miRNA_count = miRNA_expression, aff_factor = c(energy, seed_type), deg_factor = region)%>%
  update_how("Gene4", 2)%>%
  simulate(3)%>%
  activate(edges)%>%
  as_tibble()%>%
  select(Competing_name,comp_count_list,effect_list)%>%
  unnest()
```
